timesliced_random_cell_selection

### 2018-11-20 src/timesliced_random_cell_selection.Rmd
very painful depth correction...

```{r pre-install & load packages, include=FALSE}
install.packages("pacman")
library(pacman)
p_load(raster, dplyr, data.table, sp, adehabitatHR, rgdal, rworldmap)
```

@@

```{r}
mcp_unique_cell <- unique_c_sp[mcp_capelin_100, ] #this is just subsetting 
plot(mcp_capelin_100)
points(mcp_unique_cell)
```
ok now convert mcp_unique_cell back to a dataframe for further processing, and save the clipped unique cells to a .csv
```{r}
mcp_unique_cell <- as.data.frame(mcp_unique_cell)
write.csv(mcp_unique_cell, "../output/env/unique_cell_centroid_mcp")
```

ok so now I have a base list to work off... This gives you a dataset of xy points, but you need xyz (where z is based on the depth layers in the environmental data)
```{r}
unique_cell_xyz <- mcp_unique_cell
unique_cell_xyz$depthlayerno <- NA
unique_cell_xyz <- unique_cell_xyz[rep(seq_len(nrow(unique_cell_xyz)), each=50),] #this generates 50 repetitions of each row
unique_cell_xyz$depthlayerno <- rep(1:50) #assign 1:50 on a repeating cycle to the depth column. 50 is chosen because there are 50 depth layers
unique_cell_xyz$id_depth <- paste(unique_cell_xyz$glo_unique_cell, unique_cell_xyz$depthlayerno, sep="") #create a new column that is a unique glo_unique_cell & depth ID. This is to help remove cells later and randomise
```

@@

Split the data_aea into timeslices (year/month).
All you need is the year, month, cell_id, and depthlayerno
```{r}
data_aea <- mcpbase
cell_yymm <- subset(data_aea, select= c(cell_id, depthlayerno, year, month)) #create a new df with just cell_id, yr, month
cell_yymm$id_depth <- paste(cell_yymm$cell_id, cell_yymm$depthlayerno, sep="") #create a new column that is a unique cell_id & depth ID. This is to help remove cells later and randomise
split_cell_yymm <- split(cell_yymm, list(cell_yymm$year, cell_yymm$month), drop = TRUE)
```

@@

@@

Easy first one - add an id based on id_depth
```{r}
background_all$id <- background_all$id_depth
write.csv(background_all, "../output/bio/background_all.csv", row.names = FALSE)
head(background_all)
```

ok lets's get the real decimalLatitude and decimalLongitude.
Will need to plot all point, extract the lonlat coordinates

1) load the file as a spatialdatapointframe. give it the aea coordinate

```{r}
background_all <- as.data.frame(background_all)
xy <- background_all[ ,c(2,3)] #lon lat in col 1 and 2
sp_background <- SpatialPointsDataFrame(coords = xy, data = background_all, proj4string = CRS("+proj=aea +lat_1=50 +lat_2=70 +lat_0=40 +lon_0=-60 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"))
head(sp_background)
```

```{r}
plot(sp_background)
```

2) reproject to a wgs (use an existing file as a mask)

```{r}
sp_background <- spTransform(sp_background, CRS = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
plot(sp_background)
```
3) extract lonlat, add to dataframe

```{r}
sp_background <- data.frame(sp_background, long=coordinates(sp_background)[,1], lat=coordinates(sp_background)[,2])
head(sp_background)
```

remove existing decimlalLongitude and decimlalLongitude, and long and lat, and optional (duplicate columns) and change name of longitude.2 & latitude.2 columns to decimalLongitude and decimalLatitude respectively
```{r}
sp_background <- as.data.frame(sp_background) #need to go back to a 'normal' dataframe to change col names
head(sp_background) 
sp_background <- subset(sp_background, select = -c(decimalLongitude, decimalLatitude, long, lat, optional))
colnames(sp_background)[colnames(sp_background)=="longitude.2"] <- "decimalLongitude"
colnames(sp_background)[colnames(sp_background)=="latitude.2"] <- "decimalLatitude"
head(sp_background)
```

save as csv
```{r}
write.csv(sp_background, file = "../output/bio/background_all_lonlat.csv", row.names = FALSE)
```

# Add NAFO Region background point is in
NAFO Zones shapefile obtained from [NAFO](https://www.nafo.int/Data/GIS)

```{r NAFO shapefile extraction}
coordinates(sp_background) <- c("decimalLongitude", "decimalLatitude") 
nafo_zones <- readOGR(dsn=path.expand("../data/bio/nafo_zones"), layer = "nafo_zones_wgs84") #this loads the shapefile
proj4string(sp_background) <- proj4string(nafo_zones) #tells R that the occurrence data is the same projection as the shapefile
sp_background$nafo_zone <- over(sp_background, nafo_zones)$ZONE #ZONE is where the zone data is held in teh shapefiles' attributes
head(sp_background) #just to check it worked
```

Great. Lets see if any are missing a NAFO Zone. But first, write to a csv and reload as the data is now some wierd 'spatial points dataframe'...
```{r}
write.csv(sp_background, "../output/bio/background_all_lonlat_nafo.csv", row.names=FALSE)
sp_background <- as.data.frame(sp_background) 
bc_nafo_na <- subset(sp_background, is.na(sp_background$nafo_zone))
bc_nafo_na
```
ok lots don't have a NAFO region attached. This is probably the Hudson Bay issue again...

just double-check they are all gone
```{r}
background_all <- read.csv("../output/bio/background_all_lonlat_nafo2.csv", header = TRUE)
bc_nafo2_na <- subset(background_all, is.na(background_all$nafo_zone))
bc_nafo2_na
```
great none!

```{r}
head(background_all)
```
ok so some more crap has been added - the year and month both have _ appended.... change this. Also change 
- glo_unique_cell to cell_id
- longitude to longitude_meters
- latitude to latitude_meters

and remove
- optional
- x.1
- x
- id_depth

```{r}
names(background_all)[names(background_all)=="year_"] <- "year"
names(background_all)[names(background_all)=="month_"] <- "month"
names(background_all)[names(background_all)=="glo_unique_cell"] <- "cell_id"
names(background_all)[names(background_all)=="longitude"] <- "longitude_meters"
names(background_all)[names(background_all)=="latitude"] <- "latitude_meters"
background_all <- subset(background_all, select = -c(optional, X.1, X, id_depth))
head(background_all)
```

@@

>>>>

```{r pre-install & load packages, include=FALSE}
install.packages("pacman")
library(pacman)
p_load(raster, dplyr, data.table, sp, adehabitatHR, rgdal, rworldmap, ncdf4, marmap, plyr)
```

@@

```{r}
mcp_unique_cell <- unique_c_sp[mcp_capelin_100, ] #this is just subsetting 
plot(mcp_capelin_100)
points(mcp_unique_cell)
```
ok now save the clipped unique cells to a .csv
```{r}
write.csv(mcp_unique_cell, "../output/env/unique_cell_centroid_mcp")
```

#decimalLat and Lon
ok lets's get the real decimalLatitude and decimalLongitude.
Will need to plot all point, extract the lonlat coordinates

reproject to a wgs (use an existing file as a mask)

```{r}
sp_mcp_unique_cell <- spTransform(mcp_unique_cell, CRS = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
plot(sp_mcp_unique_cell)
```
extract lonlat, add to dataframe

```{r}
sp_mcp_unique_cell <- data.frame(sp_mcp_unique_cell, long=coordinates(sp_mcp_unique_cell)[,1], lat=coordinates(sp_mcp_unique_cell)[,2])
head(sp_mcp_unique_cell)
```

remove existing long, lat, optional, longitude.1, latitude.1 (duplicate columns) and change name of longitude.2 & latitude.2 columns to decimalLongitude and decimalLatitude respectively, and longitude and latitude to longitude_meters and latitude_meters ande glo_unique_cell to cell_id
```{r}
sp_mcp_unique_cell <- as.data.frame(sp_mcp_unique_cell) #need to go back to a 'normal' dataframe to change col names
head(sp_mcp_unique_cell) 
sp_mcp_unique_cell <- subset(sp_mcp_unique_cell, select = -c(long, lat, optional, longitude.1, latitude.1))
colnames(sp_mcp_unique_cell)[colnames(sp_mcp_unique_cell)=="longitude.2"] <- "decimalLongitude"
colnames(sp_mcp_unique_cell)[colnames(sp_mcp_unique_cell)=="latitude.2"] <- "decimalLatitude"
colnames(sp_mcp_unique_cell)[colnames(sp_mcp_unique_cell)=="longitude"] <- "longitude_meters"
colnames(sp_mcp_unique_cell)[colnames(sp_mcp_unique_cell)=="latitude"] <- "latitude_meters"
colnames(sp_mcp_unique_cell)[colnames(sp_mcp_unique_cell)=="glo_unique_cell"] <- "cell_id"
head(sp_mcp_unique_cell)
```

save as csv
```{r}
write.csv(sp_background, file = "../output/bio/unique_cell_centroid_lonlat.csv", row.names = FALSE)
```

# Add NAFO Region background points in
NAFO Zones shapefile obtained from [NAFO](https://www.nafo.int/Data/GIS)

```{r NAFO shapefile extraction}
coordinates(sp_mcp_unique_cell) <- c("decimalLongitude", "decimalLatitude") 
nafo_zones <- readOGR(dsn=path.expand("../data/bio/nafo_zones"), layer = "nafo_zones_wgs84") #this loads the shapefile
proj4string(sp_mcp_unique_cell) <- proj4string(nafo_zones) #tells R that the occurrence data is the same projection as the shapefile
sp_mcp_unique_cell$nafo_zone <- over(sp_mcp_unique_cell, nafo_zones)$ZONE #ZONE is where the zone data is held in teh shapefiles' attributes
head(sp_mcp_unique_cell) #just to check it worked
```

Great. Lets see if any are missing a NAFO Zone. But first, write to a csv and reload as the data is now some wierd 'spatial points dataframe'...
```{r}
write.csv(sp_mcp_unique_cell, "../output/bio/unique_cell_centroid_lonlat_nafo.csv", row.names=FALSE)
sp_mcp_unique_cell <- as.data.frame(sp_mcp_unique_cell) 
bc_nafo_na <- subset(sp_mcp_unique_cell, is.na(sp_mcp_unique_cell$nafo_zone))
bc_nafo_na
```

ok lots don't have a NAFO region attached. This is probably the Hudson Bay issue again...

just double-check they are all gone
```{r}
mcp_unique_cell <- read.csv("../output/env/unique_cell_centroid_lonlat_nafo2.csv", header = TRUE)
bc_nafo2_na <- subset(mcp_unique_cell, is.na(mcp_unique_cell$nafo_zone))
bc_nafo2_na
```

# random depth

ok so now I have a base list to work off... This gives you a dataset of xy points, but you need xyz (where z is based on the depth layers in the environmental data). This should be constrained by the presence depth range
```{r}
max_obs_depth <- max(mcpbase$depthlayerno, na.rm = TRUE) #max is 47
min_obs_depth <- min(mcpbase$depthlayerno, na.rm = TRUE) #min is 1
unique_cell_xyz <- mcp_unique_cell
unique_cell_xyz$depthlayerno <- NA
unique_cell_xyz <- unique_cell_xyz[rep(seq_len(nrow(unique_cell_xyz)), each=47),] #this generates 47  repetitions of each row 47 is chosen because there are 47 sampled depth layers
unique_cell_xyz$depthlayerno <- rep(1:47) #assign 1:47 on a repeating cycle to the depth column. 47 is chosen because there are 50 depth layers
unique_cell_xyz$id_depth <- paste(unique_cell_xyz$cell_id, unique_cell_xyz$depthlayerno, sep="") #create a new column that is a unique glo_unique_cell & depth ID. This is to help remove cells later and randomise
```

@@

Split the data_aea into timeslices (year/month).
All you need is the year, month, cell_id, and depthlayerno
```{r}
data_aea <- mcpbase
cell_yymm <- subset(data_aea, select= c(cell_id, depthlayerno, year, month)) #create a new df with just cell_id, yr, month, depthlayer no
cell_yymm$id_depth <- paste(cell_yymm$cell_id, cell_yymm$depthlayerno, sep="") #create a new column that is a unique cell_id & depth ID. This is to help remove cells later and randomise
split_cell_yymm <- split(cell_yymm, list(cell_yymm$year, cell_yymm$month), drop = TRUE)
```

@@

ok lets start by getting rid of some of the duplicated columns right now (put in by arc, various processes..)

```{r}
head(background_all)
```
remove
FID
optional

rename
longitude_ to longitude_meters
latitude_m to latitude_meters
decimalLon to decimalLongitude
decimalLat to decimalLatitude

```{r}
background_all <- subset(background_all, select = -c(optional, FID))
names(background_all)[names(background_all)=="longitude_"] <- "longitude_meters"
names(background_all)[names(background_all)=="latitude_m"] <- "latitude_meters"
names(background_all)[names(background_all)=="decimalLon"] <- "decimalLongitude"
names(background_all)[names(background_all)=="decimalLat"] <- "decimalLatitude"
head(background_all)
```

# what data to add

@@

# create unique observation ID

Easy first one - add an id based on id_depth
```{r}
names(background_all)[names(background_all)=="id_depth"] <- "id"
write.csv(background_all, "../output/bio/background_all.csv", row.names = FALSE)
head(background_all)
```

@@

the depth layers in the netCDF are literal slices. Currently I have listed the depthlayerno (which is literally which layer in the netcdf). I should also add the depth_layer name which gives the depth the slice is supposed to represent. Data is stored in depth_bin.csv

load the depth bin csv
```{r}
depthbin <- read.csv("../data/env/depth_bin.csv", header = TRUE) 
head(depthbin)
```

so basically what i want to do is, where depthbin$layer_no = background_all$depthlayerno, extract depthbin$layer_name and add to background_all$depth_layer. Can use the merge function.
First need to change depthbin col names to match background_all

```{r}
names(depthbin)[names(depthbin)=="layer_no"] <- "depthlayerno"
names(depthbin)[names(depthbin)=="layer_name"] <- "depth_layer" #this one is just so i don't have to change the column name in the background_all file later
head(depthbin)
```
```{r}
background_all <- merge(background_all, y = depthbin[ , c("depthlayerno", "depth_layer")], by = "depthlayerno", all.x = TRUE)
head(background_all)
```

just check some random rows...

```{r}
randcheck <- background_all[sample(nrow(background_all), 10), ]
randcheck
```
all good - save to .csv

```{r}
write.csv(background_all, "../output/bio/background_all_lonlat_nafo2_sciname_amo_nao_depthname.csv", row.names = FALSE)
```

# Bathymetric

bottom depth from [GEBCO](www.gebco.net)

```{r}
bat <-readGEBCO.bathy("../data/env/GEBCO_2014_2D_-70.0_34.0_-43.0_70.0.nc")
summary(bat)
```

create pretty custom color palette and then plot the bathy layer with species data
```{r}
blues <- colorRampPalette(c("lightblue", "cadetblue2", "cadetblue1", "white"))
plot(bat, n = 1, image = TRUE, bpal = blues(100))
points(background_all$decimalLongitude, background_all$decimalLatitude, pch = 20, col = "yellow") #note these are both the wgs84 images
```
lovely jubbely

and now get depth... note all the points and bathy data is wgs!

```{r}
background_all$bottom_depth <- get.depth(bat, background_all[ ,8:9], locator = FALSE)
background_all$bottom_depth <- background_all$bottom_depth[, 3] #this step is because get.bath actually produces it's own dataframe with long, lat, and depth. This is extracting the depth column and deplacing data_aea$bottom_depth with the correct data.
write.csv(background_all,"../output/bio/background_all_lonlat_nafo2_sciname_amo_nao_depthname_bath.csv", row.names = FALSE)
head(background_all)
```

# observations by cell (in total, and also by time step)

as the title says... how many observations happeend in each cell:
- in total

```{r observation by cell - total}
obs_by_cell_back <- count(background_all, "cell_id")
write.csv(obs_by_cell_back, file = "../output/bio/back_no_observations_cell.csv")
head(obs_by_cell_back)
```

- in each yyyymm
```{r}
obs_cell_yymm_back <- count(background_all, c("cell_id", "year", "month"))
write.csv(obs_cell_yymm_back, file = "../output/bio/back_no_observations_cell_yymm.csv")
head(obs_cell_yymm_back)
```

- in each yyyymm at depth
```{r}
obs_cell_yymm_depth_back <- count(background_all, c("cell_id", "year", "month", "depthlayerno"))
write.csv(obs_cell_yymm_depth_back, file = "../output/bio/back_no_observations_cell_yymm_depth.csv")
head(obs_cell_yymm_depth_back)
```

Now attach these values

total obs for the whole time period
```{r}
background_all <- merge(background_all, obs_by_cell_back, by="cell_id")
head(background_all)
```

rename freq column to something meaningful
```{r}
names(background_all)[names(background_all)=="freq"] <- "total_cell_obs_xy"
head(background_all)
```

just check some random rows...

```{r}
randcheckcell <- background_all[sample(nrow(background_all), 10), ]
randcheckcell
```
all good

time-sliced
```{r}
background_all <- merge(background_all, obs_cell_yymm_back, by.x=c("cell_id", "year", "month"), by.y=c("cell_id", "year", "month"))
names(background_all)[names(background_all)=="freq"] <- "total_cell_obs_xyt"
background_all <- merge(background_all, obs_cell_yymm_depth_back, by.x=c("cell_id", "year", "month", "depthlayerno"), by.y=c("cell_id", "year", "month", "depthlayerno"))
names(background_all)[names(background_all)=="freq"] <- "total_cell_obs_xyzt"
write.csv(background_all,"../output/bio/background_all_lonlat_nafo2_sciname_amo_nao_depthname_bath_cellobs.csv", row.names = FALSE)
head(background_all)
```

and just check some random rows...

```{r}
randcheckyymm <- background_all[sample(nrow(background_all), 10), ]
randcheckyymm
```

# attach environmental correlates

This loop runs through the netcdf files and then looks for which rows in data_aea it should extract the value to point from, and at what depth (netcDF layer)

First you need to put the background (background_all) into a spatialpointsdataframe.
```{r}
background_all2 <- subset(background_all[1:6, ])
xy <- background_all2[ ,c(7,8)] # This is to tell R where the coordinates are (in column 18 and 19). Note that the column order needs to be longitude, latitude
background_all_sp <- SpatialPointsDataFrame(coords = xy, data = background_all2, proj4string = CRS("+proj=aea +lat_1=50 +lat_2=70 +lat_0=40 +lon_0=-60 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")) # The CRS is used here is for the albers equal area projection.
head(background_all_sp)
```

```{r}
netcdf_list <- list.files("../data/env/netcdf", pattern = '*.nc', full.names = TRUE) #true means the full path is included
no_netcdf <- length(netcdf_list) #for the loop - need to know how many files to cycle through
netcdf_name <- list.files("../data/env/netcdf", pattern = '*.nc', full.names = FALSE) #false means the path is not included
aea <- raster("../output/env/aea.tif") 
for (i in 1:no_netcdf) {  
  print(netcdf_name[i]) #this just prints the name of the netCDF R is working one
  brkyr <- as.integer(sapply(strsplit(netcdf_name[i], "_"), "[[", 1)) # extracting the first part of the netcdf filename (which is the year)
  brkmth <- as.integer(sapply(strsplit(netcdf_name[i], "_"), "[[", 2)) # extracting the second part of the netcdf filename (which is the month)
  brkvar <- (sapply(strsplit(netcdf_name[i], "_"), "[[", 3)) # extracting the third part of the netcdf (inc.nc)
  temp_brick <- brick(netcdf_list[i], lvar = 4)
  temp_brick <- projectRaster(temp_brick, aea) 
    for (j in 1:nrow(background_all_sp)) {  
      de <- background_all_sp$depthlayerno[[j]]  # a variable for the observation depth layer
      yr <- (background_all_sp$year[j])  # a variable for the observation year
      mth <- (background_all_sp$month[j])  # a variable for the observation month
          if (brkyr == yr & brkmth == mth & brkvar == "temp.nc"){
              background_all_sp$temp_surface[j] <- extract(x=temp_brick[[1]], y = background_all_sp[j, ]) 
              if (is.na(de)){
                background_all_sp$temp_depth[j] <- NA
              } else  
                background_all_sp$temp_depth[j] <- extract(x=temp_brick[[de]], y = background_all_sp[j, ])
          } else if (brkyr == yr & brkmth == mth & brkvar == "salinity.nc") {
              background_all_sp$salinity_surface[j] <- extract(x=temp_brick[[1]], y = background_all_sp[j, ]) 
              if (is.na(de)){
                background_all_sp$salinity_depth[j] <- NA
              } else  
                background_all_sp$salinity_depth[j] <- extract(x=temp_brick[[de]], y = background_all_sp[j, ]) 
          } else if (brkyr == yr & brkmth == mth & brkvar == "chl.nc") {
              background_all_sp$chl_surface[j] <- extract(x=temp_brick[[1]], y = background_all_sp[j, ]) 
              if (is.na(de)){
                background_all_sp$chl_depth[j] <- NA
              } else  
                background_all_sp$chl_depth[j] <- extract(x=temp_brick[[de]], y = background_all_sp[j, ]) 
          } else if (brkyr == yr & brkmth == mth & brkvar == "o2.nc") {
              background_all_sp$o2_surface[j] <- extract(x=temp_brick[[1]], y = background_all_sp[j, ]) 
              if (is.na(de)){
                background_all_sp$o2_depth[j] <- NA
              } else  
                background_all_sp$o2_depth[j] <- extract(x=temp_brick[[de]], y = background_all_sp[j, ]) 
          } else if (brkyr == yr & brkmth == mth & brkvar == "mlp.nc") {
              background_all_sp$mlp_surface[j] <- extract(x=temp_brick[[1]], y = background_all_sp[j, ])
          } else if (brkyr == yr & brkmth == mth & brkvar == "ssh.nc") {
              background_all_sp$ssh_surface[j] <- extract(x=temp_brick[[1]], y = background_all_sp[j, ])

          }

    }
}
head(background_all_sp)
```

and just check some random rows...

```{r}
randcheck <- background_all[sample(nrow(background_all), 5), ]
randcheck
```
